---
title: "CellPlotManual"
author: "Sven E. Templer, Robert Sehlke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CellPlotManual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Description

`CellPlot` is a [GNU R](http://r-project.org) package with plot methods for the integrated visualisation of functional enrichment and expression data. It can be used to display commonly used gene ontology (GO) enrichment analysis from differential gene expression.

The development platform of this software is hosted on [github](https://github.com/dieterich-lab/CellPlot). In case there are any questions, consider using the [issues](https://github.com/dieterich-lab/CellPlot/issues) system.

### Installation

You can install the package to use it in any R session. The stable versions will be published on the comprehensive R archive network (CRAN) servers. The most recent version can be found on https://github.com/dieterich-lab/CellPlot. With the `devtools` package, it is easy to install:

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github('dieterich-lab/CellPlot', build_vignettes = TRUE)
```

### Usage

Once installed, you need to load the package with `library(CellPlot)` into the current session.
Access general help with `?CellPlot` or `?` followed by any function name from the package. 
This manual can also be viewed with `vignette("CellPlotManual")`.

## DEG and GO enrichment analysis

The following code represents a simple analysis workflow to analyse/test for differentially regulated genes 
from two conditions, perform GO enrichment analysis and merge the data to a `data.frame`. 
The example data therefore is taken from the `multtest` package, available on [Bioconductor](http://www.bioconductor.org/packages/release/bioc/html/multtest.html).
It contains microarray gene expression data from leukemia study of *Golub* et al. (1999), processed as described in *Dudoit* et al. (2002).

Differential gene expression was performed using a Student t-test to compare the two groups. 
GO annotation was done with the `annotate` and `hu6800.db` packages from Bioconductor. 
The `topGO` package was utilized to perform a GO enrichment test based on Fisher's exact test. 
Data of the DEG and GO enrichment was merged with the provided function `go.enrich`.
The code to create the data object is shown in the examples section. 
For help about the data set also read the help page (`?golubstat`) of the resulting and implemented data set.

```{r, eval=FALSE}
### Differential gene expression
### and GO enrichment analysis
### of the golub dataset:

## install dependencies
source("http://bioconductor.org/biocLite.R")
biocLite("hu6800.db")
biocLite("multtest")
biocLite("topGO")

## load dependencies
library(CellPlot)
library(plyr)
library(multtest)
library(annotate)
library(hu6800.db)
library(topGO)

## data import
data(golub)

## differential gene expression testing
golub.test <- function (e, i, n) {
  i <- as.logical(i)
  a <- split(e[,i],seq(nrow(e)))
  b <- split(e[,!i],seq(nrow(e)))
  fun.map <- function (a, b) {
    t <- t.test(a, b)
    t$fc <- t$estimate[1]/t$estimate[2]
    suppressWarnings(t$fc <- log2(t$fc))
    data.frame(
      mean.aml = t$estimate[1],
      mean.all = t$estimate[2],
      log2fc  = t$fc,
      p = t$p.value,
      row.names = NULL)
  }
  s <- Map(fun.map, a, b)
  s <- do.call(rbind, s)
  s$p.adj <- p.adjust(s$p)
  s <- data.frame(n, s, stringsAsFactors = FALSE)
  s <- subset(s, !is.na(log2fc))
  return(s)
}

degdata <- golub.test(golub, golub.cl, data.frame(
  gene = golub.gnames[,3], index = as.integer(golub.gnames[,1]),
  stringsAsFactors = FALSE))

## gene ontology annotation
goterms <- select(
  hu6800.db, degdata$gene,
  c("PROBEID","ALIAS","GO","ENSEMBL","ENTREZID"))
goterms <- dlply(
  subset(goterms, ONTOLOGY == "BP"), "PROBEID",
  function (x) unique(x$GO))

## topGO object
godata <- new(
  "topGOdata", ontology = "BP", description = 'golub',
  allGenes = setNames(degdata$p.adj, degdata$gene),
  geneSelectionFun = function (allScore) { allScore <= 0.05 },
  annotationFun = annFUN.gene2GO, gene2GO = goterms)

## gene ontology enrichment
#golubstat <- go.enrich(godata, degdata, p.max = 1)
golubstat <- go.enrich(godata, degdata)
```

A table like `degdata` from this example might also be obtained from DEG analysis with common tools 
as used in the Tuxedo pipeline (tophat/cuffdiff) and imported with `read.table`. 
The resulting data object is available with our package. You can load it with the `data` function.
The columns are described in the help page, and the data structure can be investigated with the `str` and `head` functions, e.g.:

```{r}
library(CellPlot)
data("golubstat")
str(head(subset(golubstat, !duplicated(genes))))
```

## Visualization of GO enrichment

For the examples, we use a subset of the GO enrichment analysis of the `golubstat` data. Create the subset with:

```{r}
x <- subset(golubstat, p<=.05 & significant>4 & !duplicated(genes))
x <- head(x, 10)
```

### Cell plot

The function `cell.plot` plots a horizontal barchart of strictly positive values in x. 
For each entry, a vector of additional values needs to be provided in a list. 
The additional values are plotted as cells subdividing the length of the corresponding bar. 
A typical usage scenario is plotting the enrichment of Gene Ontology terms, 
with individual cells reflecting the differential regulation of the constituent genes.

```{r, fig.width=7, fig.height=6}
cell.plot( x = setNames(x$loge, x$term), cells = x$deg.log2fc, x.bound = 2, 
           cell.bounds = c(-6,6), space = 0.15, main ="", y.mar = c(0.1,0), 
           x.mar = c(.47, 0), key.n = 7, cex = 1.6, cell.outer = 3 )
```

### Sym plot

The function `sym.plot` plots a split barchart, showing the proportions of 
two mutually exclusive sets in relation to a set containing them both. 
E.g., Gene Ontology terms, showing the proportions of differentially down-regulated 
and up-regulated annotated genes from a perturbation experiment. 
The color of the central column elements maps to the value provided in x (e.g. GO term enrichment). 
Associated genes may be provided as a list of vectors of expression values, 
same as for cell.plot(), or as separate vectors x.up and x.down, 
providing the numbers of up- and down-regulated genes in the same order as x.

```{r, fig.width=7, fig.height=6}
sym.plot( ticksize = 5, x = setNames(x$loge, x$term), cells = x$deg.log2fc, 
          x.annotated = x$annotated, y.mar = c(0.1,0), x.mar = c(.47, 0), key.n = 7, 
          key.lab = "GO Term enrichment", cex = 1.6, axis.cex = 0.8, group.cex = 0.7 )
```

### Arc plot

```{r, fig.width=7, fig.height=6}
x$deg.up <- lapply(Map(setNames, x$deg.log2fc, x$genes), function (i) { i[i>0] })
x$deg.down <- lapply(Map(setNames, x$deg.log2fc, x$genes), function (i) { i[i<0] })
arc.plot( x = setNames(x$loge, x$term), up.list = x$deg.up, down.list = x$deg.down, x.mar=c(1,0.5))
```

### Visual parameters

Both functions provide a range of parameters to tweak the visual appearance of the plots.

* **Font sizes** of the various elements can be adjusted using the `cex` parameters.
* **Internal margins** are adjusted through `x.mar` and `y.mar`. Used in particular to accomodate the label text to the left-hand side.
* **Absolute scaling** can be achieved through the `bar.scale` parameter, which acts as a multiplier for a predetermined average bar height. This is meant to be used in conjunction with a graphics device of constant size, to ensure visual consistency among multiple plots that vary in the number of terms displayed.
* **Data ranges** for all output may be fixed:
    * `elem.bounds` Require the cardinality of a term or significant subset to be in a specific range, e.g. `c(10,100)` -- between 10 and 100 genes.
    * `x.bound` Upper limit of the value displayed on the x-axis. For `sym.plot` this is must be a value between 0 and 100. The lower limit is always 0.
    * `cell.bounds` (only `cell.plot`) Sets the colour scale for cell data to a fixed range. If the plot contains values outside of that range, an indicator is added to the colour legend. This is particularly useful in the event of a few extreme outliers.
    * `mid.bounds` (only `sym.plot`) Similar to `cell.bounds`, this sets the colour range of the central column cells.
    * `sym` (only `cell.plot`), when set to `TRUE`, makes the scale of cell values symmetrical.
* **Colour schemes** may be controlled through character vectors provided to the following parameters. Internally, the R native function `ColorRampPalette()` is used.
    * `cell.col` (only `cell.plot`) Specify three valid colour names for cell data visualisation (lower extreme, mid point, upper extreme).
    * `mid.col` (only `sym.plot`) Specify two valid colour names to define the range of the central column cell colours. 
* **Colour keys** can be adjusted in their resolution (number of boxes) using the `key.n` parameter.
* **Gridlines** may be hidden via the `gridlines` parameter.

